---
title: "Tutorial ggplot2"
author: Dr. med. Ramon Saccilotto, Departement of Clinical Research,<br> University
  Hospital Basel, Switzerland, June 1st 2015
output:
  ioslides_presentation:
    self_contained: no
    widescreen: yes
---

## The ggplot2 package

In essence ggplot2 is a collection of functions that will help you to create 
maintainable and publication ready plots in an efficient manner.

The package was developed by [Hadley Wickham, assistant professor of statistics 
at Rice University, New Zealand](http://had.co.nz) and is currently maintained 
by the author himself and a number of volunteers on github.

Descriptions and examples of almost all of the packages functions can be found
in the fantastic [online documentation](http://docs.ggplot2.org/current/).


## The ggplot2 package

If you are using ggplot on a regular basis, I highly recommend to read at 
least one of the following books for further information:

- H.Wickham, ggplot2, Use R
- Paul Murrell, R Graphics, Second Edition


## Philosophy

> «ggplot2 is an R package for producing statistical, or data, graphics, 
but it is unlike most other graphics packages because it has a deep 
underlying grammar.»

> «This grammar, based on the Grammar of Graphics (Wilkinson, 2005), is composed 
of a set of independent components that can be composed in many different ways. 
[..]»

> «Plots can be built up iteratively and edited later.»

> «A carefuly chosen set of defaults means that most of the time you can 
produce a publication-quality graphic in seconds, but if you do have speical 
formatting requirements, a comprehensive theming system makes it easy to do 
what you want. [..]»

## Philosophy

> «ggplot2 is designed to work in a layered fashion, starting with a layer 
showing the raw data then adding layers of annotation and statistical 
summaries. [..]»

> «ggplot2 is a plotting system for R, based on the grammar of graphics, which 
tries to take the good parts of base and latticegraphics and none of the bad 
parts.»

> «It takes care of many of the fiddly details that make plotting a hassle 
(like drawing legends) as well as providing a powerful model of graphics that 
makes it easy to produce complex multi-layered graphics.»


# Tutorial

---

> This tutorial is intended to give you a brief overview of the amazing 
ggplot2 package (currently at version 1.0.1). 

> After completion you should be able to quickly create a variety
of different plots and have the necessary understanding to adapt the plots
to your specific needs.

> Please note however, that we will not cover all the functionality of the 
ggplot2 package. 

> More information and examples of almost all functions and plot types can be 
found in the [online documentation](http://docs.ggplot2.org/current/)

---

Your fist step should be to install and load the ggplot2 package.

```{r install ggplot2, eval=FALSE}
# first we need to install the ggplot and some supporting libraries 
# (skip this step if the library is already loaded)
install.packages("ggplot2")

```

```{r load ggplot2-package, message=FALSE, warning=FALSE}
# we will require the ggplot2 package for our graphics
# note: there are some additional useful packages such as plyr, 
#  reshape2 and scales which you may find useful
require("ggplot2")

```


## Some data included in the package

```{r data frames to use, cache=TRUE}
# prices of 50'000 sparkly round cut diamonds
head(diamonds)

# motor trend car road tests
head(mtcars)

# vapor pressure of mercury at certain temperatures
head(pressure)

```


## How to use ggplot

- In contrast to the regular plot() function, ggplot is based on the grid-
package for drawing all plots (as is lattice)
- Therefore, ggplot is not «really» compatible with the default plot-functions

- Plots are drawn in layers that are stacked on top of each other

- To create a plot we either use the **qplot()** or **ggplot()** function


## qplot vs ggplot

- In general, qplot() is used for standard plots without much configuration
- The ggplot() syntax allows more specific plot definitions, but seems to be a 
little bit more complicated at first

> However, once you understand the underlying principle, the ggplot syntax is 
easy to comprehend and very well suited for more complex plots.

___

```{r qplot vs ggplot, cache=TRUE}
# histogram with qplot
qplot(clarity, data=diamonds, fill=cut, geom="bar")

# histogram with ggplot -> same output
ggplot(diamonds, aes(clarity, fill=cut)) + geom_bar()

```


## How to use qplot to quickly create some nice graphs

**Important note: ggplot (and qplot for that matter) always expect the data 
to be in a data.frame**

The syntax for qplot is `qplot(«x-axis», «y-axis», data=«data.frame», ..)`

The ggplot2 package has an extendable `fortify` method which can be used
to convert R-objects to data-frames for plotting. 

Some useful fortify methods are already availabe in the package. 
More can be found in the `ggfortify` package on github.

---

```{r data transformation, cache=TRUE}
# quickly create a scatterplot of our data
qplot(wt, mpg, data=mtcars)

# data can be transformed with functions
qplot(log(wt), mpg-10, data=mtcars)

```

---
```{r variable mapping, cache=TRUE}
# plots can be further refined by using additional parameters
# note: we are mapping the variable «qsec» to a color
qplot(wt, mpg, data=mtcars, color=qsec)

```


## Mapping variables into the plot

Plot-attributes such as color, point, shape, etc. are called aesthetics.

With qplot(), assigning a variable to an aesthetic will map the values of the 
variable into the value-space of the aesthetic.

Note: Both the american «color» and british «colour» are
supported in most cases.

---
```{r mapping vs setting, cache=TRUE}
# color and colour will work for most cases
qplot(wt, mpg, data=mtcars, color=qsec)

# note: colour instead of color
qplot(wt, mpg, data=mtcars, colour=qsec)

# note: in this example ggplot is trying to map the size of a point 
#  to a scale of [10] (which is probably not as intended)
qplot(wt, mpg, data=mtcars, color=qsec, size=10)

# use the I() function «as is» to set aesthetics instead of mapping
qplot(wt, mpg, data=mtcars, color=qsec, size=I(10))
```

```{r alpha transparency, cache=TRUE}
# side note: it is possible to use alpha-blending for overlapping elements
qplot(wt, mpg, data=mtcars, color=qsec, size=I(10), alpha=qsec)

# note: alpha-opacity is set between 0 (transparent) and 1 (opaque)
qplot(wt, mpg, data=mtcars, color=qsec, size=I(10), alpha=I(0.5))		

```

--- 
```{r discrete vs continuous scale, cache=TRUE}
# we take a closer look at the variable cyl from the dataset mtcars
# note: the variable is stored as a continuous number not as a factor
head(mtcars)
summary(mtcars$cyl)
table(mtcars$cyl)

# regular numeric variables will be mapped to a continuous scale
qplot(wt, mpg, data=mtcars, color=cyl)

# factored variables will be displayed with a discrete scale
qplot(wt, mpg, data=mtcars, color=factor(cyl))

```

## Some further qplot examples

```{r auto plot selection, cache=TRUE}
# ggplot will try to guess the «correct» plot for your data
qplot(wt, mpg, data=mtcars)
qplot(factor(cyl), data=mtcars)

```

---
```{r setting the type of plot, cache=TRUE}
# a specific type of plot can be set with the attribute geom=«type»
qplot(wt, mpg, data=mtcars, geom="point")
qplot(wt, mpg, data=mtcars, geom="line")

```

---
```{r multiple plot types, cache=TRUE}
# plot-types can be combined
qplot(wt, mpg, data=mtcars, geom=c("line", "point"))

# note: problem if only size of points should be increased
qplot(wt, mpg, data=mtcars, geom=c("line", "point"), size=I(2))

# pro-tipp: resort to ggplot syntax (more on that later)
qplot(wt, mpg, data=mtcars) + geom_line() + geom_point(size=4)

```

---
```{r flipping the plot by 90°, cache=TRUE}
# a plot can be flipped by 90°
# note: coord_flip() will rotate the plot after calculation of 
#  any summary statistics (i.e. smoothers or alike)
qplot(factor(cyl), data=mtcars)
qplot(factor(cyl), data=mtcars) + coord_flip()

```

---
```{r fill vs color, cache=TRUE}
# difference between fill/color bars
qplot(factor(cyl), data=mtcars, fill=factor(cyl))
qplot(factor(cyl), data=mtcars, color=factor(cyl))

```

---
```{r position adjustments, cache=TRUE}
# use different position properties for bars (stacked, dodged, fill, identity)
head(diamonds)
qplot(clarity, data=diamonds, geom="bar", fill=cut, position="stack")
qplot(clarity, data=diamonds, geom="bar", fill=cut, position="dodge")
qplot(clarity, data=diamonds, geom="bar", fill=cut, position="fill")
qplot(clarity, data=diamonds, geom="bar", fill=cut, position="identity")

```

## The ggplot syntax | (or building plots layer by layer)

The ggplot syntax is used to build a plot layer by layer. Usually, the
following steps are involved

- Defining the data to be used in the plot with `ggplot(«data.frame»)`
- Specifing the visual representation of the data with geoms, i.e. `geom_point()`
or `geom_line()`
- Specifing the features or aestethics to represent the values in the plot with
`aes()`
- Optionally modifying scales, labels or adding additional layers

> Note: The underlying data is always the same for all layers - although 
there is a workaround that you should only rarely use


## Building a plot

```{r ggplot2 syntax, cache=TRUE, error=TRUE, eval=-2}
# we are going to use some pressure data
head(pressure)

# nothing happens if we only define our data
ggplot(pressure)
```

---
```{r ggplot2 syntax (2), cache=TRUE}
# but we can quickly add a representation
# note: the aes() function is used for variable mapping
ggplot(pressure) + geom_point(aes(x=temperature, y=pressure))

# as x and y are used so often, we can leave it of
# note: for later maintenance it is usually better to specify it
ggplot(pressure) + geom_point(aes(temperature, pressure))

# note: you can access the previously created plot with «last_plot()»
last_plot()

```

---
```{r defining aesthetics, cache=TRUE}
# specify a value allocation outside of the aes() function
# if an aestetic should be set to a specific value
ggplot(pressure) + geom_point(aes(temperature, pressure), size=4)

# aesthetics can also be defined separately
ggplot(pressure) + aes(temperature, pressure) + geom_point(size=4)

```

---
```{r dealing with overplotting, cache=TRUE}
# create some normal distributed test data
tmp <- data.frame(x=rnorm(4000), y=rnorm(4000))
p.myplot <- ggplot(tmp, aes(x,y))

# default plotting
p.myplot + geom_point(color="red")

# plotting using hollow circles
p.myplot + geom_point(shape=1, color="red")

# plotting using pixels
p.myplot + geom_point(shape=".", color="red")

# plotting using alpha transparency
# note: requires the scales package (included with ggplot2)
p.myplot + geom_point(color=scales::alpha("red", 1/2))
p.myplot + geom_point(color=scales::alpha("red", 1/6))

```

---
```{r ggplot objects, cache=TRUE}
# ggplot will actually return an object that can be modified
# note: the object can also be saved for later use with save()
#   saving a plot or layer definitions will also include the plot data
p.myplot <- ggplot(pressure)

# summary information about the plot
summary(p.myplot)

# adding some additional layers
p.myplot <- p.myplot + aes(temperature, pressure) + geom_point(size=4)
summary(p.myplot)

# the plot can be printed by just calling the object or using print()
p.myplot
print(p.myplot)

````

```{r updating the underlying data, cache=TRUE}
# the underlying data is saved within the ggplot-object. modifications of 
# the data will not alter the plot if the plot-code is not rerun.
# there is however a special syntax to run the plot with updated data
pressure2 <- data.frame(
	"temperature"=pressure$temperature, "pressure"=log(pressure$pressure))

# print the plot with updated data
p.myplot %+% pressure2

```

---
```{r exporting ggplot graphics, cache=TRUE, eval=FALSE}
# a plot can be exported using ggsave
# note: the respective rendering device needs to be installed
ggsave(file="testplot.pdf", plot=p.myplot, width=10, height=5)
ggsave(file="testplot.svg", plot=p.myplot, width=10, height=5)
ggsave(file="testplot.png", plot=p.myplot, dpi=72, width=10, height=5)

```

---
```{r order of layers, cache=TRUE}
# let's define a base plot and aesthetic-mapping
p.myplot <- ggplot(pressure) + aes(x=temperature, y=pressure)

# using multiple layers
p.myplot + 
	geom_point(color="purple3", size=6) + 
	geom_line(color="steelblue2", size=2)

# the order of the layers does mather 
# (each new layer is drawn on top of the previous)
p.myplot + 
	geom_line(color="steelblue", size=2) +
	geom_point(color="purple3", size=6)

```

---
```{r aesthetics and settings for multiple layers, cache=TRUE}
# aesthetics defined in the base layer will be used for all layers
# note: setting attributes to a value will not apply it to other layers
ggplot(pressure, aes(x=temperature, y=pressure), color="red") + 
	geom_line(size=4, alpha=0.3) + 
	geom_point(size=4)

```

---
```{r ggplot layer arguments, cache=TRUE}	
# the actual arguments to map variables is mapping=«aes()» and 
# geom_params=«list()» to set variables respectively
ggplot(pressure) + 
	geom_point(
		mapping=aes(x=temperature, y=pressure, color=factor(temperature)),
		geom_params=list(size=4, shape=18)
	)

```


## Various syntax options

> qplot()
> ggplot()
> geom_«type»()
> layer()

---
```{r various syntax options, cache=TRUE}
# it is possible to mix qplot and ggplot
qplot(temperature, pressure, data=pressure, geom="line", lty=I("dashed")) + 
	geom_point(size=4)

# there is also a different syntax with layer()
ggplot(pressure, aes(temperature, pressure)) +
	layer(geom="line", mapping=aes(color=temperature), size=4) +
	layer(geom="point", size=4, color="purple3")
	
# let's use some additional data in the plot
# note: the scales of the different datasets are unified
ggplot(pressure) +
	aes(x=temperature, y=pressure) +
	layer(geom="line", mapping=aes(color=temperature), size=4) +
	geom_point(data=mtcars, aes(hp, disp), color="purple3", size=3)

```

## Scales

Scales are required to give the plot reader a sense of reference and thus
encompass the ideas of both axes and legends on plots.

ggplot2 will usually automatically chose appropriate scales and display 
legends if necessary. 

It is however, very easy to override or modify the default values.

The scale syntax is scale_«attribute»_«optional subspecification»()
i.e. `scale_x_continuous()` or `scale_x_discrete()`

---

```{r scales, cache=TRUE}
# setup our plot with default scales
p.myplot <- ggplot(pressure) + 
	aes(temperature, pressure, color=factor(temperature)) +
	geom_point(size=4)

# scales can be limited to a certain range
p.myplot
p.myplot + scale_x_continuous("Temperature", limits=c(200, 400))

# scales that are used as axes will take the name as axis label
p.myplot + 
	scale_color_discrete(name="Temperature \nin C°") + 
	scale_y_continuous(name="Air pressure at sea level")

# legends can also be removed (if not important to understand the plot)
p.myplot + scale_color_discrete(guide="none")

```

---
```{r naming scales and legends, cache=TRUE}
# setup a different plot
p.myplot <- ggplot(diamonds, aes(cut, fill=color)) + geom_bar()
p.myplot

# the axis can be renamed using two different methods
p.myplot + xlab("Diamond Cut")
p.myplot + scale_x_discrete(name="Diamond Cut Description")
p.myplot + scale_y_continuous(name="Number of Diamonds")

# names of legends can also be set
p.myplot + scale_fill_discrete(name="Diamond Color")

```

---
```{r color scales, cache=TRUE}
# using some custom colors
# note: brewer colors were created for good readable maps and often provide
#  a good alternative to the standard colors. to see all available brewer
#  palettes use «RColorBrewer::display.brewer.all()»
p.myplot + scale_fill_grey()
p.myplot + scale_fill_hue()
p.myplot + scale_fill_brewer()
p.myplot + scale_fill_brewer(type="seq", palette="3")
p.myplot + scale_fill_brewer(palette="Paired")

```

---
```{r custom color scales, cache=TRUE}
# using a custom color palette with specified order
# note: color values should be specified as hex or color names
p.myplot + aes(fill=cut) + scale_fill_manual(
	values = c("#7fc6bc","#083642","#b1df01",
						 "#cdef9c","#466b5d", "#744db5", "#ccb2e8"))

# using predefinded colors for specific values
# note: values that are not present in the data will not be shown
p.myplot + aes(fill=cut) + scale_fill_manual(
	values = c("Fair"="#083642", "Good"="#466b5d", 
						 "Very Good"="#7fc6bc","Premium"="#cdef9c", 
						 "Ideal"="#b1df01", "Not specified"="#ffffff"))

# removing values from the legend and custom labelling of values
# note: you must specify colors for all existing values
p.myplot + scale_fill_manual(
	name="Colors",
	values = c("D"="#083642", "E"="#466b5d", "F"="#7fc6bc","G"="#cdef9c", 
						 "H"="#b1df01", "I"="#ababab", "J"="#ececec"),
	breaks = c("D", "E", "F"),
	labels = c("E"="Dark Green", "D"="Esmerald", "F"="Wood"))

```

---
```{r modify legends, cache=TRUE}
# legends can also be styled using guides
# note: guides can be defined once and be easily applied to multiple plots
p.mylegend <- guide_legend(
		title="Color of the \nDiamond",
		title.position="top",
		direction="horizontal",
		label.position="top",
		label.hjust=0.5,
		label.vjust=0.5,
		ncol=2,
		byrow=TRUE,
	)

# apply some styling to the legend
p.myplot + guides(fill = p.mylegend)
p.myplot + scale_fill_discrete(guide=p.mylegend)

```


---
```{r handling transparency in legends, cache=TRUE}
# handling problems with alpha transparency
p.myplot + aes(alpha=color)

# remove the alpha transparency for the legend
p.myplot + aes(alpha=color) + 
	guides(fill = guide_legend( override.aes=list(alpha=1) ))

```

---
```{r limits vs zooming in, cache=TRUE}
# limiting scales will remove all points that are outside of the scale
# note: be careful, this is not the same as just focusing on a graph region
p.myplot + scale_y_continuous(limits=c(0,15000))

# to focus on a specific region, the coord_cartesian() function
#  should be used with the specified limits
p.myplot + coord_cartesian(ylim=c(0,15000))

```

## Statistical transformations

Some graphical representations do not use the raw data directly, but 
perform a statistical transformation - i.e. binning.

Several transformations are included in the ggplot2 package and can be
called with the stat_«transformation»() functions.

Examples are `stat_bin`, `stat_boxplot`, `stat_qq`, `stat_unique`,
`stat_smooth`, `stat_summary` and more

---

```{r histogram statistics, cache=TRUE}
# histograms will use stat_bin to calculate number of items per bin
ggplot(mtcars) + aes(qsec) + geom_histogram(binwidth=0.5)
ggplot(mtcars) + aes(qsec) + geom_histogram(binwidth=1)

```

---
```{r regressions/smoothers, cache=TRUE}
# define a base plot to illustrate smoothed lines
p.myplot <- ggplot(mtcars) + aes(x=disp, y=mpg) + geom_point(size=4)
p.myplot

# draw a smooth line (local regression function) through the points
# note: the default smoothing function is loess
p.myplot + geom_line(stat="smooth")

# using the smooth geom with standard deviation
p.myplot + geom_smooth()

# fit the regression closer to the data with span=«0-1»
p.myplot + geom_smooth(span=0.4)
p.myplot + geom_smooth(span=1)

# turning off the confidence interval
# note: the attribute level can be used to set ci-level
p.myplot + geom_smooth(se=FALSE)

```

---
```{r regression methods, cache=TRUE}
# using a different method for smoothing (i.e. linear modelling)
p.myplot + geom_smooth(method="lm")

# using a cutom formular for fitting
library(splines)
p.myplot + geom_smooth(method="lm", formula = y ~ ns(x,5))

# be careful when flippling a plot
# note: details on transformations on the following slide
p.myplot + geom_smooth()
p.myplot + geom_smooth() + coord_flip()
p.myplot + aes(x=mpg, y=disp) + geom_smooth()

```

## Scale and coordinate transformations

Values can be transformed either before or after any stats-functions are 
applied.

Use scale transformations to apply a transformation before any stats-function
are applied.

Use coordinate transformations to apply a transformation after all stats-
functions are applied (note: `coord_flip()` is a coordinate transformation).

---
```{r scale and coord transformations, cache=TRUE}
# define a base plot to illustrate transformation 
p.myplot <- ggplot(mtcars) + aes(x=disp, y=mpg) + geom_point(size=4) +
	geom_smooth(method="lm", se=FALSE)

# take a look at linear regression plot
p.myplot

# apply a logarhithmic transformation 
p.myplot + scale_x_continuous(trans="log", name="log(disp)")

# apply a log-transformation on the y-axis, add a linear regression and
# transform the display of the scale back with exponentation
p.myplot + scale_x_continuous(trans="log") + 
	coord_trans(x="exp") +
	xlab("exp(log(disp)) = disp")

# adjust the y-scale breaks to match our original non transformed plot
p.myplot + scale_x_continuous(trans="log", breaks=seq(100,400,100)) + 
	coord_trans(x="exp") +
	xlab("exp(log(disp)) = disp")
	
```

## Subgroups | Groups and Facets

The ggplot2 package provides two interesting functionalities to look at
subgroups in your data.

The `group` aesthetic is useful if there are only two or three groups and 
there is a summary statistic that should be calculated per group and 
displayed in one chart.

Facettings on the other hand is useful to split the data into different groups
that are displayed next to each other.

---
```{r grouping of variables, cache=TRUE}
# split data to create frequency polygon for each subgroup
qplot(clarity, data=diamonds, geom="bar", fill=cut, position="dodge")
qplot(clarity, data=diamonds, geom="freqpoly", group=cut, color=cut, position="identity")

# split the data by a variable and calculate a regression for each group
ggplot(mtcars, aes(x=disp, y=mpg, color=factor(am))) + geom_point(size=4) +
	geom_smooth(aes(group=factor(am)), method="lm", se=FALSE, lty="dashed")

```

---
```{r facets, cache=TRUE}
# use facets to split the data
p.myplot <- ggplot(mtcars) +
	aes(x=disp, y=mpg, color=factor(am)) + 
	geom_point(size=4) +
	geom_smooth(method="lm", se=FALSE, lty="dashed")

# facet_wrap will wrap the specified panels
p.myplot + facet_wrap(~ am, nrow=1)
p.myplot + facet_wrap(~ am, ncol=1)

# per default the scales of the different panels will match
# it is however possible to use adaptive panes
# note: more options can be found in the documentation
p.myplot + facet_wrap(~ am, nrow=1, scales="free")

```

---
```{r facet-grid syntax, cache=TRUE}
# facet_grid can be used to split by two variables
p.myplot + facet_grid(cyl ~ am)

# it is even possible to add margin calculations
p.myplot + facet_grid(cyl ~ am, margins=TRUE)

```

## Annotations

Annotations can be added as separate layers that will not influence 
the scales or legends

```{r annotations, cache=FALSE}
# setup plot to illustrate annotations
p.myplot = ggplot(mtcars, aes(x = wt, y = mpg))

# plot without annotations
p.myplot + geom_point(size=4, color="purple3")

# a plot with some simple annotations
p.myplot + 
	annotate("rect", 
					 fill="lightsteelblue", alpha=0.4,
					 xmin=3, xmax=4, ymin=12, ymax=20.5) + 
	annotate("segment", 
					 size = 1, color="steelblue", 
					 arrow = grid::arrow(length=grid::unit(1, "char")),
					 x=4.73, y=30.5, xend=3.8, yend=21) +
	annotate("text", label="A custom region", 
					 x=4.32, y=31.2, hjust=0, vjust=0, color="steelblue", size=6) + 
	geom_point(size=4, color="purple3")  

```

---

Unfortunately, it is currently not possible to paint textures or 
patterns as backgrounds with the ggplot2 package.

However, we may easily define our own pattern creating function.

```{r drawing a striped pattern, cache=TRUE}
# we create a function, that will calculate the coordinates for stripes that
# are contained to the given rect coordinates
# note: this involves some trigonometry and is outside 
#   the scoope of this tutorial
stripesInRect <- function(angle=45, distance=0.5, xmin=0, xmax=10, ymin=0, ymax=10) {
	# this function will calculate a data.frame of vectors for a
	# stripped background in a rectangular area

	# convert angle from degree to radians
	radians <- (pi / 180) * angle

	# calculate the tangens
	tangens <- tan(radians)

	# calculate height und width of the clippling box
	height <- ymax - ymin
	width <- xmax - xmin

	# calculate the horizontal distance of the lines
	horizontalDistance = distance / tangens

	# calculate the difference of start-y to end-y for full width
	verticalDifference <- tangens * width

	# steps for the height and width
	stepsHeight = seq(from = ymin, to = ymax, by = distance)
	stepsWidth = seq(from = xmin, to = xmax, by = horizontalDistance)

	# initialize a data frame of coordinates
	# note: distance is used for distance of lines when cutting
	#   through the side of the box
	# note: we have to remove the first step from the widthsteps
	#  to avoid a duplicated start line
	data <- data.frame(
		"x1" = c(rep(xmin, times = length(stepsHeight)), stepsWidth[-1]),
		"y1" = c(stepsHeight, rep(ymin, length(stepsWidth))[-1] ))

	# define a function to calculate the endpoints
	calculateEndpoint <- function(x1, y1) {

		# calculate the maximal available width for the x range
		availableWidthRange <- xmax - x1

		if (availableWidthRange >= width) {
			# calculation of lines that start from the left side

			# calculate the maximal available height for the y range
			availableHeightRange <- ymax - y1

			# we are done if the vertical-side fits into the rect
			if (availableHeightRange >= verticalDifference) {
				return(c(
					"x2" = xmax,
					"y2" = y1 + verticalDifference))
			}

			# otherwise we have to adapt to the available height
			horizontalDifference <- availableHeightRange / tangens

			return(c(
				"x2" = x1 + horizontalDifference,
				"y2" = ymax))

		} else {
			# calculation of lines that start from the bottom side

			# calculate the vertical difference
			verticalDifference <- availableWidthRange * tangens

			return(c(
				"x2" = xmax,
				"y2" = y1 + verticalDifference
				))

		}

	}

	# calculate the endpoints
	endpoints <- mapply(calculateEndpoint, data$x1, data$y1)

	# extract the endpoint coordinates
	data$x2 <- endpoints[1,]
	data$y2 <- endpoints[2,]

	return(data)
}


# calculate the pattern coordinates for our plot
pattern <- stripesInRect(angle=80, distance=0.25, 
		xmin=3, xmax=4, ymin=12, ymax=20.5)

# create the plot with a striped background for the annotation
# note: annotation aesthetics are not mapped but will be processed as vectors
p.myplot + 
	annotate("segment", 
					 size = 0.5, color="deeppink", alpha=0.25,
					 x=pattern$x1, y = pattern$y1, 
					 xend = pattern$x2, yend = pattern$y2) + 
	annotate("segment", 
					 size = 1, color="deeppink", 
					 arrow = grid::arrow(length=grid::unit(1, "char")),
					 x=4.73, y=30.5, xend=3.8, yend=21) +
	annotate("text", label="A custom region", 
					 x=4.32, y=31.2, hjust=0, vjust=0, color="deeppink", size=6) + 
	geom_point(size=4, color="purple3")  
	
```


## Themes

The ggplot2 package separates the data-part of the plot from the non-data
part. 

The collection of graphical parameters used to control non-data elements - such
as the background-color, font-sizes, .. - are controlled in so called themes

It is possible to create custom themes that can be applied to multiple plots
easily.

Use the function `theme_set(«theme»)` to set a global theme for all plots.

---
```{r theme example, cache=TRUE}
# define our plot to illustrate theming
p.myplot <- ggplot(pressure) + 
	aes(temperature, pressure, color=factor(temperature)) +
	geom_point(size=4)

# plotting using the default theme
p.myplot

# plotting using a black & white theme
# note: the theme does not change the aesthetics controlled by data
p.myplot + theme_bw()

```

---
```{r custom theme modifications, cache=TRUE}
# modifiying specific elements of a theme
# note: more options can be found in the documentation
#  theme modifications may require some understanding of the grid-package
p.myplot + theme(
	legend.position="top", 
	legend.margin=grid::unit(1, "cm"))

# legends usually need some further specific adjustments
p.myplot + theme(
	legend.position="bottom", 
	legend.margin=grid::unit(1, "cm")) +
 guides(
 	color=guide_legend("Temperature", nrow=2, 
 										 title.position="top", byrow=TRUE))

```

```{r barplot like lattice, cache=TRUE}
# use combination of geoms and specific stat for bin calculation
# note: values from stat-calculations can be accessed via ..«parameter»..
ggplot(mtcars) + aes(x=factor(gear)) + 
	layer(
		stat = "bin",
		geom = "linerange", 
		geom_params = list(ymin=0, size=0.5, color="blue"), 
		mapping = aes(ymax=..count..)) +
	layer(
		stat = "bin",
		geom = "point",
		geom_params = list(size=3, color="blue")) +
	layer(
		stat = "bin",
		geom = "text",
		geom_params = list(vjust=-0.8, color="blue"),
		mapping = aes(label=..count..)) +
	coord_flip() + theme_bw()

```

---
```{r a resusable lattice-like barplot function, cache=TRUE}
# we can also define the configuration in a custom function
latticebars <- function(color = "blue") {
	layer1 <- layer(
		geom = "linerange", stat = "bin",
		mapping = aes(ymax=..count..),
		geom_params = list(ymin=0, size=0.5, color=color))
		
	layer2 <- layer(
		geom = "point", stat = "bin", 
		geom_params = list(size=3, color=color))
	
	layer3 <- layer(
		 geom = "text", stat = "bin",
		mapping = aes(label=..count..),
		geom_params = list(vjust=-0.8, color=color))
		
	# note: ggplot2 elements can also be combined by creating
	#  a list of the separate components. +-symbol might
	#  throw an error if used inside of a function
	return(list(layer1,layer2,layer3, coord_flip(), theme_bw()))
}

# create a lattice like barplot with default color
ggplot(mtcars) + aes(x=factor(gear)) + 
	latticebars()

# easily change the color of the plot
ggplot(mtcars) + aes(x=factor(gear)) + 
	latticebars("red") + 
	xlab("Type of Gear\n") + ylab("\nNumber of Items")
	
```


## Multiple plots in one graphic

Ggplot2 is based on the grid-package. Therefore, it is possible to fit
multiple plots on the same page using the `viewports` functionality.

Further information about using viewports is available in the grid-package
documentation.

```{r require grid package, message=FALSE, warning=FALSE}
# we are going to need the grid package
require("grid")
```

```{r convenience functions for multi-plots, cache=TRUE, message=FALSE, warning=FALSE}
# convenience function to create multi-plot setup (nrow, ncol)
vp.setup <- function(x,y){
	# create a new layout with grid
	grid.newpage()
	
	# define viewports and assign it to grid layout 
	pushViewport(viewport(layout = grid.layout(x,y)))
}

# convenience function to easily access layout (row, col)
vp.layout <- function(x,y){ 
	viewport(layout.pos.row=x, layout.pos.col=y)
}

```

---
```{r multiple plots in on graphic, cache=TRUE, message=FALSE, warning=FALSE}
# define three plots to be displayed together
p.a <- qplot(mpg, wt, data=mtcars, geom="point") + theme_bw()
p.b <- qplot(mpg, wt, data=mtcars, geom="bar", stat="identity")
p.c <- qplot(mpg, wt, data=mtcars, geom="step")

# setup amulti plot layout with grid (2x2 fields)
vp.setup(2,2)

# plot all graphics into our layout
print(p.a, vp=vp.layout(1, 1:2)) 
print(p.b, vp=vp.layout(2, 1)) 
print(p.c, vp=vp.layout(2, 2))

```


